---
title: "Forecast hospital census from hubverse admission forecasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecast hospital census from hubverse admission forecasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Public-health forecasting hubs (FluSight, COVIDhub, RSVhub) publish
**admission** forecasts. Hospital operations plan around the
**census**, the number of patients in hospital on a given day.

`censcast` transforms hubverse admission forecasts into census
forecasts by convolving them with a length-of-stay (LOS) distribution.
The result lands back in the hubverse quantile-forecast schema.

## The model

If $P(\text{LOS} > d)$ is the probability that a patient is still in
hospital $d$ time-steps after admission, then today's census is past
admissions weighted by their probability of still being here:

$$
\text{census}(t) \;=\; \sum_{s \le t} \text{admissions}(s) \cdot
P(\text{LOS} > t - s).
$$

Apply this convolution to each quantile of a hubverse admission
forecast and you get a hubverse census forecast: same schema in,
same schema out.

## Pipeline

You bring three things, all in standard hubverse shape:

* an admission quantile forecast — `hubData::collect_hub()`;
* observed admissions covering the `max_stay` window before the
  forecast starts — `hubData::connect_target_timeseries()`. Census
  on day one depends on patients admitted before it, so the
  convolution needs that history to start from;
* a LOS distribution — `los_spec()`. Bring values from literature,
  hospital records, or a separate fitting pipeline.

```{r}
library(censcast)
library(hubData)
library(tidyverse)

max_stay <- 3 # weeks
start_date  <- Sys.Date() - 100
burnin_date <- start_date - max_stay * 7   # earliest history date needed
los <- los_spec("negbin", mu = 0.6, k = 1.7, max_stay = max_stay)

admission_fc <- connect_hub(s3_bucket("cdcepi-flusight-forecast-hub")) |>
  filter(
    target         == "wk inc flu hosp",
    output_type    == "quantile",
    model_id       == "FluSight-ensemble",
    reference_date >= start_date
  ) |>
  collect_hub()

admission_hist <- connect_target_timeseries(
  s3_bucket("cdcepi-flusight-forecast-hub")
) |>
  filter(
    target          == "wk inc flu hosp",
    target_end_date >= burnin_date
  ) |>
  collect()

census_fc <- admission_fc |>
  fcast_census(los = los, history = admission_hist)

plot_fan(census_fc, location = "US")
```

## Comparing LOS assumptions

Without ground-truth census, LOS is the dominant assumption. The
principled response is to run several specs and compare. Tag each run
with a label and bind:

```{r}
library(purrr)

los_specs <- list(
  "short (mu=0.3w)" = los_spec("negbin", mu = 0.3, k = 1.7, max_stay = max_stay),
  "med (mu=1w)"     = los_spec("negbin", mu = 1,   k = 1.7, max_stay = max_stay),
  "long (mu=3w)"    = los_spec("negbin", mu = 3,   k = 1.7, max_stay = max_stay)
)

census_runs <- imap(los_specs, \(los, name) {
  admission_fc |>
    fcast_census(los = los, history = admission_hist) |>
    mutate(los = name)
}) |>
  bind_rows()

plot_fan(census_runs, location = "US") +
  facet_wrap(~ los, ncol = 1, scales = "free_y")
```

`plot_fan()` treats any column it doesn't recognise as a trajectory
grouper, so the user-added `los` column needs no configuration.


## Scoring

Hubverse hubs publish admissions truth, not census truth, so to
score you bring your own. It needs the standard hubverse target-data
columns: `target_end_date`, `target`, `location`, `observation`.

The demo below fakes one by convolving observed admissions with a
"true" LOS. In real use, replace `observed_census` with your
observed-census source.

```{r}
true_los <- los_spec("negbin", mu = 0.6, k = 1.7, max_stay = max_stay)

observed_census <- admission_hist |>
  arrange(location, target_end_date) |>
  group_by(location) |>
  mutate(observation = stats::convolve(observation, rev(true_los),
                                       type = "open")[seq_along(observation)]) |>
  ungroup()

scores_by_los <- imap(los_specs, \(los, name) {
  admission_fc |>
    fcast_census(los = los, history = admission_hist) |>
    score_census(truth = observed_census) |>
    mutate(los = name)
}) |> bind_rows()

scores_by_los |>
  summarise(wis = mean(wis), .by = c(los, horizon)) |>
  ggplot(aes(horizon, wis, color = los)) +
  geom_line(linewidth = 0.7) + 
  geom_point() +
  theme_bw()

scores_by_los |>
  summarise(across(c(overprediction, underprediction, dispersion), mean),
            .by = c(los, horizon)) |>
  pivot_longer(-c(los, horizon), names_to = "component", values_to = "wis") |>
  ggplot(aes(horizon, wis, fill = component)) +
  geom_col() +
  facet_wrap(~ los) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Horizon (weeks)", y = "Mean WIS") +
  theme_bw()
```


## API

| Function         | Role |
|------------------|------|
| `los_spec()`     | Build a LOS survival vector. |
| `fcast_census()` | Convolve hubverse admissions × LOS → hubverse census. |
| `plot_fan()`     | Quantile fan chart, returns a `ggplot`. |
| `score_census()` | Score against observed census via `scoringutils`. |


