---
title: "Forecast hospital census from forecasting hubs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecast hospital census from forecasting hubs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  message = FALSE,
  warning = FALSE
)
```

Public health forecasting hubs (FluSight, COVIDhub, RSVhub) publish
**admission** forecasts. Hospital operations plan around the
**census**: the number of patients in hospital on a given day.
`censcast` turns one into the other by convolving admissions with a
length of stay (LOS) distribution:

$$
\text{census}(t) \;=\; \sum_{d \ge 0} \text{admissions}(t - d) \cdot
P(\text{LOS} > d).
$$

```{r setup}
library(censcast)
library(dplyr)
library(purrr)
library(ggplot2)
```

## What the hubverse gives you

Two tibbles in standard hubverse shape, both shipped with the package
as toy data for this vignette:

* `admission_history`: observed weekly admissions, as you would pull
  from a hub's target time series (`hubData::connect_target_timeseries()`).
* `admission_forecast`: the next 4 weeks of admission quantile
  forecasts, as you would pull from a hub
  (`hubData::connect_hub() |> collect_hub()`).

Overlaying the two on one axis shows what a hub publishes: history in
black, forecast fan in blue.

```{r inputs}
plot_fan(admission_forecast, truth = admission_history) +
  labs(y = "Weekly admissions")
```

`censcast` gives two ways to turn this into a *census* forecast,
depending on whether you have past census data.

## 1. With past census data: `fit_los()`

Hubs publish admissions truth, not census truth, so the census series
has to come from your own records. The package ships a toy version,
`census_history`, generated by convolving `admission_history` with a
known LOS, so it pairs row for row with the admission series.

`fit_los()` recovers the LOS by deconvolving the paired admission and
census time series:

```{r fit-route}
fit <- fit_los(
  data = inner_join(
    admission_history |> select(target_end_date, admissions = observation),
    census_history |> select(target_end_date, census = observation),
    by = "target_end_date"
  ),
  family = "negbin",
  max_stay = 8
)
fit
```

`plot(fit)` shows the recovered survival curve:

```{r fit-survival, fig.height = 3}
plot(fit)
```

Drop the fit into `fcast_census()` to get a census forecast, then
overlay the observed census to check the alignment:

```{r fit-forecast}
census_fit <- fcast_census(admission_forecast, fit, admission_history)
plot_fan(census_fit, truth = census_history) +
  labs(y = "Weekly census")
```

## 2. Without past census data: `spec_los()`

If no census data is available, set LOS from literature priors. Running
several priors makes the assumption visible. Short stays put weight on
recent admissions; long stays smooth across many weeks:

```{r priors, fig.height = 7}
los_priors <- list(
  "short (mu=0.5w)" = spec_los("negbin", mu = 0.5, k = 1.7, max_stay = 8),
  "medium (mu=2w)" = spec_los("negbin", mu = 2, k = 1.7, max_stay = 8),
  "long (mu=4w)" = spec_los("negbin", mu = 4, k = 1.7, max_stay = 8)
)

census_priors <- imap(los_priors, \(s, name) {
  fcast_census(admission_forecast, s, admission_history) |>
    mutate(los = name)
}) |>
  bind_rows()

# Replicate truth across facets so each panel gets the black line.
truth_faceted <- imap(los_priors, \(s, name) {
  mutate(census_history, los = name)
}) |>
  bind_rows()

plot_fan(census_priors, truth = truth_faceted) +
  facet_wrap(~los, ncol = 1) +
  labs(y = "Weekly census")
```

`plot_fan()` treats any unrecognised column as a trajectory grouper,
so the user added `los` column needs no configuration.

## 3. Scoring

`score_census()` returns weighted interval score (WIS) per quantile
via `scoringutils`. Pair each forecast with the truth and compare mean
WIS across LOS priors:

```{r scoring}
imap(los_priors, \(s, name) {
  fcast_census(admission_forecast, s, admission_history) |>
    score_census(census_truth = census_history) |>
    mutate(los = name)
}) |>
  bind_rows() |>
  summarise(mean_wis = mean(wis), .by = los) |>
  arrange(mean_wis)
```

The "medium" prior matches the LOS used to generate `census_history`
(`negbin(mu = 2, k = 1.7)`), so it wins. In a real workflow, lean on
`fit_los()` when census data exists and use `spec_los()` priors as a
sensitivity check.

## Plugging in a real hub

The toy datasets above are the only difference between this vignette
and a production workflow. Swap them for live hub data:

```{r real-hub, eval = FALSE}
library(hubData)

admission_forecast <- connect_hub(s3_bucket("cdcepi-flusight-forecast-hub")) |>
  dplyr::filter(
    target == "wk inc flu hosp",
    output_type == "quantile",
    model_id == "FluSight-ensemble"
  ) |>
  collect_hub()

admission_history <- connect_target_timeseries(
  s3_bucket("cdcepi-flusight-forecast-hub")
) |>
  dplyr::filter(target == "wk inc flu hosp") |>
  dplyr::collect()
```

`census_history` (or its real equivalent) you supply from your own
records.

## API

| Function         | Role |
|------------------|------|
| `spec_los()`     | LOS survival vector from priors. |
| `fit_los()`      | LOS survival vector fit from `(admissions, census)`. |
| `fcast_census()` | Convolve admissions × LOS → census, hubverse shape in/out. |
| `plot_fan()`     | Quantile fan chart, returns a `ggplot`. |
| `score_census()` | Score census forecast against truth via `scoringutils`. |
